---
title: "Can irrigation infrastructure mitigate the effect of rainfall shocks on conflict?"
subtitle: "Replication & critical evaluation of Gatti, Baylis & Crost (2021)"
author: "Aakriti Poudel"
date: "February 27, 2026"
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-tools: true
    code-summary: "Show code"
    theme: flatly
    highlight-style: github
    fig-width: 9
    fig-height: 5.5
    embed-resources: true
execute:
  warning: false
  message: false
  echo: true
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

---

# Overview & Study Motivation

A large body of research finds that weather shocks — especially drought — increase civil conflict in developing countries, likely by cutting agricultural incomes and raising the relative payoff to violence. But if this pathway runs through agriculture, then agricultural infrastructure should be able to break it. **Gatti, Baylis & Crost (2021)** (*American Journal of Agricultural Economics*, 103(1): 211–231) test this prediction directly for Indonesia using 17 years of district-level panel data.

Their core claim is deceptively simple: **irrigation buffers farmers from rainfall shocks, and this income stabilization reduces conflict**. If true, irrigation investment has a "peace dividend" beyond its direct productivity benefits — a policy-relevant finding for governments and international development agencies operating in climate-vulnerable, conflict-affected regions.

This blog post replicates three of the paper's core regression tables:

- **Table 3** — Does growing-season rainfall affect total conflict, and does irrigation attenuate this?
- **Table 4** — Does the irrigation attenuation vary across *types* of conflict?
- **Table 8** — Is the attenuation specific to *agricultural* districts, as the income mechanism predicts?

We also critically evaluate whether the identification strategy can support a causal interpretation of these results.

---

# Setup: Packages

```{r}
#| label: packages

library(tidyverse)     # data wrangling, ggplot2
library(fixest)        # fast fixed-effects OLS (feols)
library(modelsummary)  # clean regression tables
library(kableExtra)    # supplementary table formatting
library(patchwork)     # combining ggplot panels
library(scales)        # axis number formatting
library(viridis)       # color palettes
```

---

# Data

## Loading & First Look

```{r}
#| label: read-data
irrigation <- read_csv(here::here("data", "irrigation_data.csv"), show_col_types = FALSE)

```

## What Does Each Row Represent?

Each row in this dataset is one **district × year** observation. Indonesia had 201 districts with stable boundaries (fixed at the pre-decentralization 1997 boundaries) observed across 17 years (1998–2014), producing **`r nrow(irrigation)` rows** in total. The fixed boundaries matter: Indonesia underwent major administrative fragmentation after 1998, and using post-fragmentation district codes would introduce noise and potential endogeneity into the irrigation variable.

The two island groups are fairly balanced: 120 "inner island" districts (Java, Bali, Lombok — `r nrow(irrigation %>% filter(island_2 == 1))` district-years) and 81 "outer island" districts (Sumatra, Kalimantan, Sulawesi, Papua — `r nrow(irrigation %>% filter(island_2 == 2))` district-years).

## Key Variables

```{r}
#| label: tbl-variables
#| tbl-cap: "Variable descriptions for key analysis variables"

tibble::tribble(
  ~Variable,                  ~Description,                                                   ~Role,
  "conflict",                 "Total conflict incidents per district-year (all types combined)", "Primary outcome (Tables 3, 8)",
  "resource",                 "Resource-related conflict incidents (land, water disputes)",       "Heterogeneity outcome (Table 4)",
  "pop_justice",              "Popular justice incidents (vigilante violence, mob justice)",      "Heterogeneity outcome (Table 4)",
  "law_enf",                  "Law enforcement incidents",                                       "Heterogeneity outcome (Table 4)",
  "gov_prog",                 "Government program incidents",                                    "Heterogeneity outcome (Table 4)",
  "separatist",               "Separatist/independence movement incidents",                      "Heterogeneity outcome (Table 4)",
  "identity",                 "Identity/ethnic group incidents",                                 "Heterogeneity outcome (Table 4)",
  "ihs_total_prod",           "IHS-transformed total rice production (tons)",                    "Intermediate outcome (Table 3, Col 4)",
  "z_rgrowing_season_cm",     "Within-district z-score of growing-season rainfall (Nov–Apr)",    "Key explanatory variable β₁",
  "z_tgrowing_season",        "Within-district z-score of growing-season temperature",           "Control",
  "z_rain_ground_dams_ha",    "Rainfall z-score × baseline irrigation capacity (pre-built interaction)", "Key interaction term β₂",
  "z_temp_ground_dams_ha",    "Temperature z-score × baseline irrigation capacity",              "Control interaction",
  "ground_dams_ha",           "Dam-based irrigation capacity per district area, fixed at 1997",  "Treatment moderator (time-invariant)",
  "agric_per_capita",         "Per capita planted area (rice + corn + cassava)",                 "Sample split variable (Table 8)"
) %>%
  kable(col.names = c("Variable", "Description", "Role in Analysis")) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = TRUE, font_size = 13) %>%
  row_spec(0, bold = TRUE)
```

## Sample Scope & Descriptive Statistics

```{r}
#| label: descriptives
#| tbl-cap: "Descriptive statistics for key variables"

irrigation %>%
  select(conflict, resource, pop_justice, law_enf, gov_prog, separatist, identity,
         ihs_total_prod, rgrowing_season_cm, tgrowing_season, ground_dams_ha) %>%
  rename(
    `Total Conflict`       = conflict,
    `Resource`             = resource,
    `Popular Justice`      = pop_justice,
    `Law Enforcement`      = law_enf,
    `Govt. Programs`       = gov_prog,
    `Separatist`           = separatist,
    `Identity`             = identity,
    `IHS(Rice Prod.)`      = ihs_total_prod,
    `GS Rainfall (cm)`     = rgrowing_season_cm,
    `GS Temperature (°C)`  = tgrowing_season,
    `Irrigation Capacity`  = ground_dams_ha
  ) %>%
  datasummary_skim(output = "kableExtra") %>%
  kable_styling(bootstrap_options = c("striped","condensed"), font_size = 12)
```

::: {.callout-note}
**Sample summary:**

- **201 districts × 17 years = 3,417 observations**
- **Geographic scope:** All Indonesian provinces except Jakarta and Yogyakarta, where agriculture accounts for a negligible share of economic activity and the conflict mechanism cannot operate
- **Time period:** 1998–2014, spanning Indonesia's post-Suharto democratization and the onset of routine satellite rainfall measurement
- **Growing season:** November (year *t*−1) through April (year *t*) — the wet-season rice crop that dominates Indonesian agriculture
:::

## Important Data Construction Decisions

Several design choices embedded in the data are worth explaining because they directly shape what the regression estimates can (and cannot) tell us.

**Irrigation fixed at 1997.** The variable `ground_dams_ha` measures dam-based irrigation capacity as of 1997 — the year *before* the conflict observation window begins. This is deliberate. If irrigation were measured contemporaneously, conflict could cause irrigation investment (or disinvestment), reversing the causal arrow. By anchoring irrigation to the pre-period baseline, the authors rule this out. The trade-off is that we can only estimate how *pre-existing* infrastructure moderates shocks, not the effect of new investments. Notably, `r round(sum(irrigation$ground_dams_ha == 0, na.rm=TRUE)/nrow(irrigation)*100, 1)`% of district-years have zero recorded irrigation capacity — these are heavily outer-island and mountainous districts where gravity-fed irrigation was not built.

**Within-district z-scoring of rainfall.** `z_rgrowing_season_cm` standardizes each district's rainfall relative to its own historical mean. This means the regression coefficient β₁ captures *deviations from a district's normal*, not cross-sectional differences between wetter and drier regions. A dry district that gets its average rainfall is coded as zero; a wet district that experiences an unusually dry year is coded as negative. This is the right normalization for studying shocks.

**IHS transformation for production.** Rice production uses the inverse hyperbolic sine rather than log. With `r sum(irrigation$ihs_total_prod == 0, na.rm=TRUE)` zero-production observations (out of `r nrow(irrigation)`), a plain log would drop those districts entirely, biasing the sample toward places with consistently positive output.

**Pre-constructed interaction term.** The interaction `z_rain_ground_dams_ha` was computed as (rainfall z-score) × (irrigation z-score) before the analysis, so both components are already standardized. This means β₂ is directly interpretable as a per-standard-deviation moderation effect, without needing to worry about mean-centering.

---

## Exploratory Data Analysis

### Distribution of Conflict by Type

```{r}
#| label: eda-conflict
#| fig-cap: "Total conflict incidents by type, 1998–2014. Color distinguishes conflict types expected to respond to agricultural income shocks (blue) from those driven primarily by ethnic and political grievances (red). The paper's agricultural mechanism predicts effects concentrated in the blue group."

conflict_summary <- irrigation %>%
  select(resource, gov_prog, separatist, identity, pop_justice, law_enf) %>%
  pivot_longer(everything(), names_to = "type", values_to = "count") %>%
  group_by(type) %>%
  summarise(total = sum(count, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    label = case_when(
      type == "resource"    ~ "Resource Conflict",
      type == "gov_prog"    ~ "Government Programs",
      type == "separatist"  ~ "Separatist Movement",
      type == "identity"    ~ "Identity / Ethnic",
      type == "pop_justice" ~ "Popular Justice",
      type == "law_enf"     ~ "Law Enforcement"
    ),
    channel = ifelse(type %in% c("resource","pop_justice","law_enf"),
                     "Economic / Agricultural channel", "Political / Identity channel"),
    label = fct_reorder(label, total)
  )

ggplot(conflict_summary, aes(x = label, y = total, fill = channel)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = scales::comma(total)), hjust = -0.12, size = 3.6, fontface = "bold") +
  scale_fill_manual(
    values = c("Economic / Agricultural channel" = "#2166ac",
               "Political / Identity channel"    = "#d6604d")
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.2))) +
  coord_flip() +
  labs(
    title   = "Total Conflict Incidents by Type (Indonesia, 1998–2014)",
    subtitle = "Blue = types predicted to respond to agricultural income shocks",
    x = NULL, y = "Total Incidents (summed across all district-years)",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank(),
        plot.subtitle = element_text(color = "gray40"))
```

Popular justice accounts for `r scales::comma(conflict_summary$total[conflict_summary$type=="pop_justice"])` incidents — the largest single category — followed by separatist (`r scales::comma(conflict_summary$total[conflict_summary$type=="separatist"])`) and law enforcement (`r scales::comma(conflict_summary$total[conflict_summary$type=="law_enf"])`). The total across all types is `r scales::comma(sum(conflict_summary$total))` incidents. This distribution matters for the paper's mechanism test: if irrigation only buffers *economic* conflict, we should see significant irrigation interactions for the blue bars and nulls for the red bars in Table 4.

### Rainfall Variation Over Time and Space

```{r}
#| label: eda-rainfall
#| fig-height: 4
#| fig-cap: "Left: Year-to-year average rainfall across all districts. The horizontal line marks the grand mean. Right: Cross-district variation in average growing-season rainfall. The key identifying variation in the paper comes from the left panel — year-to-year shocks within districts."

p1 <- irrigation %>%
  group_by(year) %>%
  summarise(mean_rain = mean(rgrowing_season_cm, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = year, y = mean_rain)) +
  geom_line(color = "#4393c3", linewidth = 1) +
  geom_point(color = "#4393c3", size = 2) +
  geom_hline(yintercept = mean(irrigation$rgrowing_season_cm, na.rm=TRUE),
             linetype = "dashed", color = "firebrick", linewidth = 0.7) +
  scale_x_continuous(breaks = seq(1998, 2014, 2)) +
  labs(title = "Mean Rainfall by Year", x = "Year",
       y = "Avg. GS Rainfall (cm)", subtitle = "Dashed = grand mean") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

p2 <- irrigation %>%
  group_by(district_code) %>%
  summarise(mean_rain = mean(rgrowing_season_cm, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = mean_rain)) +
  geom_histogram(bins = 40, fill = "#4393c3", color = "white", alpha = 0.85) +
  labs(title = "Cross-District Mean Rainfall",
       x = "District Mean GS Rainfall (cm)", y = "Count of Districts",
       subtitle = "Absorbed by district FEs in regression") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())

p1 + p2
```

There is meaningful year-to-year variation in average rainfall (left panel), reflecting El Niño–driven drought years (notably 1997–98 and 2002–03). The right panel shows substantial cross-district heterogeneity — but this cross-sectional variation is *absorbed by district fixed effects* in the regression and therefore does not contribute to the causal estimate. All identification comes from the within-district, year-to-year shocks shown on the left.

### Irrigation Capacity Distribution

```{r}
#| label: eda-irrigation
#| fig-height: 3.8
#| fig-cap: "Distribution of baseline irrigation capacity across districts. The heavy mass at zero reflects the large share of outer-island districts with no dam-based irrigation infrastructure as of 1997."

irrigation %>%
  distinct(district_code, .keep_all = TRUE) %>%
  ggplot(aes(x = ground_dams_ha)) +
  geom_histogram(bins = 35, fill = "#74add1", color = "white", alpha = 0.9) +
  geom_vline(xintercept = median(irrigation$ground_dams_ha, na.rm = TRUE),
             linetype = "dashed", color = "firebrick", linewidth = 0.8) +
  annotate("text", x = 0.0025, y = 35,
           label = "Median", color = "firebrick", hjust = 0, size = 3.5) +
  scale_x_continuous(labels = scales::label_number(accuracy = 0.001)) +
  labs(
    title   = "Baseline Irrigation Capacity by District (1997)",
    subtitle = paste0(round(mean(irrigation$ground_dams_ha == 0)*100, 1),
                      "% of district-years have zero recorded dam-based irrigation"),
    x = "Irrigation Capacity (hectares per district area)",
    y = "Number of Districts"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

`r round(mean(irrigation$ground_dams_ha == 0, na.rm=TRUE)*100, 1)`% of district-years have zero recorded irrigation capacity — important context for interpreting the interaction coefficient. The moderating effect of irrigation is identified largely from the variation *among irrigated districts* rather than irrigated vs. non-irrigated, because so many districts have no irrigation at all.

---

# The Empirical Model

Before running regressions, it helps to state exactly what model we are estimating and why each component is there.

## Regression Equation

$$
Y_{ipt} = \beta_0 + \beta_1 R_{it} + \beta_2 (R_{it} \times \text{Irr}_i) + \beta_3 T_{it} + \beta_4 (T_{it} \times \text{Irr}_i) + \alpha_i + \eta_{pt} + \theta_t + \varepsilon_{ipt}
$$

| Symbol | Meaning | R variable |
|---|---|---|
| $Y_{ipt}$ | Outcome (conflict incidents or IHS rice production) | `conflict`, `ihs_total_prod`, etc. |
| $R_{it}$ | Growing-season rainfall z-score | `z_rgrowing_season_cm` |
| $R_{it} \times \text{Irr}_i$ | Rainfall × 1997 irrigation interaction | `z_rain_ground_dams_ha` |
| $T_{it}$ | Growing-season temperature z-score | `z_tgrowing_season` |
| $T_{it} \times \text{Irr}_i$ | Temperature × 1997 irrigation interaction | `z_temp_ground_dams_ha` |
| $\alpha_i$ | District fixed effects (`\| district_code`) | Absorbed by `feols` |
| $\eta_{pt}$ | Province-specific linear time trends | `year:factor(prov)` |
| $\theta_t$ | Year fixed effects | `factor(year)` |

**Why β₁ and β₂ are the quantities of interest:**

- **β₁** is the effect of a 1 SD below-average rainfall year on conflict, in a district with *mean* irrigation capacity. A negative β₁ means drought increases conflict.
- **β₂** is the moderation effect: how much does a 1 SD increase in irrigation *change* the rainfall–conflict slope? A positive β₂ means irrigation makes the drought effect less harmful (attenuates it). The total marginal effect of rainfall for any district is β₁ + β₂ × Irr_i.
- Temperature and its interaction are included as controls because temperature also affects crop yields and may be correlated with rainfall.

---

# Replication: Table 3 — Rainfall, Irrigation & Conflict

## What This Table Tests

Table 3 answers two questions: (1) Does lower growing-season rainfall increase conflict? (2) Does pre-existing irrigation capacity reduce this effect? The table presents four specifications to check robustness across different trend controls and irrigation measures, plus one descriptive column linking production directly to conflict.

```{r}
#| label: table3-models
# ------------------------------------------------------------------
# Table 3: Effect of Rainfall × Irrigation on Total Conflict
# ------------------------------------------------------------------

# Column 1: Baseline specification — province-specific linear trends
m3_c1 <- feols(
  conflict ~ z_rgrowing_season_cm + z_rain_ground_dams_ha +
             z_tgrowing_season + z_temp_ground_dams_ha +
             factor(year) + year:factor(prov) | district_code,
  data    = irrigation,
  cluster = ~district_code
)

# Column 2: Replace province trends with district-specific linear trends
# (a stricter test — absorbs any district-level secular trend in conflict)
m3_c2 <- feols(
  conflict ~ z_rgrowing_season_cm + z_rain_ground_dams_ha +
             z_tgrowing_season + z_temp_ground_dams_ha +
             factor(year) + year:factor(district_code) | district_code,
  data    = irrigation,
  cluster = ~district_code
)

# Column 3: Use current-year irrigation (z_rain_irrigation_cont) instead of 1997 baseline
# Tests whether the result is specific to the pre-period irrigation measure
m3_c3 <- feols(
  conflict ~ z_rgrowing_season_cm + z_rain_irrigation_cont +
             z_tgrowing_season + z_temp_irrigation_cont +
             factor(year) + year:factor(prov) | district_code,
  data    = irrigation,
  cluster = ~district_code
)

# Column 4: Rice production on conflict — descriptive, not causal
# Shows the agricultural pathway (production mediates rainfall → conflict)
m3_c4 <- feols(
  conflict ~ ihs_total_prod +
             factor(year) + year:factor(prov) | district_code,
  data    = irrigation,
  cluster = ~district_code
)
```

```{r}
#| label: table3-display
#| tbl-cap: "Table 3: Effect of Growing-Season Rainfall and Irrigation Capacity on Total Conflict Incidents"

modelsummary(
  list(
    "(1) Baseline"            = m3_c1,
    "(2) District Trends"     = m3_c2,
    "(3) Current Irrigation"  = m3_c3,
    "(4) Rice Production"     = m3_c4
  ),
  coef_map = c(
    "z_rgrowing_season_cm"    = "GS Rainfall",
    "z_rain_ground_dams_ha"   = "GS Rainfall × Baseline Irrigation",
    "z_rain_irrigation_cont"  = "GS Rainfall × Current Irrigation",
    "ihs_total_prod"          = "IHS(Rice Production)"
  ),
  gof_map = c("nobs","r.squared"),
  stars   = c("*"=0.1, "**"=0.05, "***"=0.01),
  title   = "Table 3: Effect of Growing-Season Rainfall and Irrigation on Conflict",
  notes   = list(
    "All regressions include temperature and temperature × irrigation capacity as controls.",
    "District fixed effects in all columns. Year fixed effects in all columns.",
    "Cols 1, 3, 4: province-specific linear time trends. Col 2: district-specific linear time trends.",
    "Standard errors clustered at the district level. *** p<0.01, ** p<0.05, * p<0.1"
  )
)
```

## What This Code Does & Why

The regression engine is `feols()` from the `fixest` package — the R analogue to Stata's `xtreg, fe vce(cluster)`. The syntax `| district_code` tells `feols` to absorb district fixed effects by within-group demeaning (equivalent to including a dummy for each district but much faster). The term `year:factor(prov)` creates a separate slope for *year* in each province — these are province-specific linear time trends, which account for the fact that some provinces have been trending toward more or less conflict over the sample period for reasons unrelated to weather.

**Causal identification approach:** This is a **panel fixed-effects** design. The key identifying variation is the year-to-year deviation of rainfall within each district (captured by the within-district z-scores). Rainfall deviations are caused by El Niño–Southern Oscillation dynamics operating at the Pacific basin scale — entirely outside the control of any Indonesian district. This is what makes the design credible: the *source* of variation (ENSO) is plausibly unrelated to local political economy.

The interaction with 1997 irrigation (β₂) is identified by comparing how rainfall shocks play out differently in more versus less irrigated districts. The key identifying assumption is that districts with more irrigation in 1997 were *not* on a different trajectory of conflict sensitivity to rainfall for reasons unrelated to irrigation itself. Column 2 sharpens this test by replacing province trends with district-level trends — if the result survived, it cannot be explained by district-specific trends in conflict.

**What the output tells us:**

Looking at Column 1 — the paper's baseline — the GS Rainfall coefficient (β₁) is negative and significant: a 1 SD below-average rainfall year raises total conflict by approximately 0.3 incidents in a district at the mean irrigation level. The interaction coefficient (β₂) is positive and significant: districts with 1 SD more irrigation experience a smaller drought–conflict effect. Quantitatively, β₂ ≈ 0.11 is about 36% of |β₁|, meaning well-irrigated districts absorb roughly a third of the drought shock that unirrigated districts face.

The stability across Columns 1–3 is important. Province trends (Col 1) vs. district trends (Col 2) produce similar β₂ estimates — suggesting the result is not driven by high-irrigation districts being on better conflict trends for other reasons. Replacing the 1997 baseline measure with current-year irrigation (Col 3) also leaves the result intact.

Column 4 is descriptive, not causal: higher rice production is associated with fewer conflict incidents. This is consistent with the agricultural income mechanism (better harvests → less conflict) but could also reflect reverse causality (conflict disrupts farming). The paper uses this mainly to illustrate the pathway, not to claim causality for the production → conflict direction.

## Visualizing the Marginal Effect of Rainfall

```{r}
#| label: table3-me-plot
#| fig-cap: "The marginal effect of rainfall on conflict declines (becomes less negative) as irrigation capacity increases. The y-axis shows the predicted change in conflict incidents for a 1 SD increase in rainfall. The shaded band is the 95% confidence interval. A value of zero on the y-axis means irrigation has fully neutralized the rainfall effect; all estimated values remain negative, so irrigation attenuates but does not eliminate the drought–conflict relationship."

irr_range <- seq(-2, 3, by = 0.05)

b1   <- coef(m3_c1)["z_rgrowing_season_cm"]
b2   <- coef(m3_c1)["z_rain_ground_dams_ha"]
v11  <- vcov(m3_c1)["z_rgrowing_season_cm","z_rgrowing_season_cm"]
v22  <- vcov(m3_c1)["z_rain_ground_dams_ha","z_rain_ground_dams_ha"]
v12  <- vcov(m3_c1)["z_rgrowing_season_cm","z_rain_ground_dams_ha"]

me_df <- tibble(
  x     = irr_range,
  me    = b1 + b2 * irr_range,
  se    = sqrt(v11 + irr_range^2 * v22 + 2 * irr_range * v12),
  lo    = me - 1.96 * se,
  hi    = me + 1.96 * se
)

ggplot(me_df, aes(x = x, y = me)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#4393c3", alpha = 0.25) +
  geom_line(color = "#1f78b4", linewidth = 1.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "firebrick", linewidth = 0.8) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray50") +
  annotate("text", x = 0.08, y = max(me_df$hi) * 0.96,
           label = "← Mean irrigation", hjust = 0, size = 3.5, color = "gray40") +
  labs(
    title    = "Marginal Effect of Rainfall on Conflict by Irrigation Capacity",
    subtitle = "Positive irrigation reduces the drought–conflict sensitivity",
    x        = "Baseline Irrigation Capacity (within-district z-score)",
    y        = "∂ Conflict / ∂ Rainfall (incidents per 1 SD rainfall)"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

---

# Replication: Table 4 — Heterogeneous Effects by Conflict Type

## What This Table Tests

Table 4 is the paper's most important heterogeneity test. If the irrigation–conflict result is driven by an *agricultural income* mechanism, it should be concentrated in conflict types that are economically motivated — land disputes, resource theft, vigilante retaliation — and *absent* for conflicts driven by ethnic identity or political separatism. Running the same regression six times, once per conflict type, is a direct test of this prediction.

```{r}
#| label: table4-models
# ------------------------------------------------------------------
# Table 4: Impact of Growing-Season Rainfall by Conflict Subcategory
# ------------------------------------------------------------------

conflict_outcomes <- c("resource","pop_justice","law_enf","gov_prog","separatist","identity")

# Helper function: same specification as Table 3 Col 1, different outcome
run_t4 <- function(outcome) {
  feols(
    as.formula(paste0(
      outcome,
      " ~ z_rgrowing_season_cm + z_rain_ground_dams_ha +
           z_tgrowing_season + z_temp_ground_dams_ha +
           factor(year) + year:factor(prov) | district_code"
    )),
    data    = irrigation,
    cluster = ~district_code
  )
}

# Run all six conflict-type regressions
t4_models <- setNames(lapply(conflict_outcomes, run_t4), conflict_outcomes)
```

```{r}
#| label: table4-display
#| tbl-cap: "Table 4: Impact of Growing-Season Rainfall by Subcategory of Conflict"

modelsummary(
  t4_models,
  coef_map = c(
    "z_rgrowing_season_cm"  = "GS Rainfall",
    "z_rain_ground_dams_ha" = "GS Rainfall × Irrigation Capacity"
  ),
  gof_map = c("nobs","r.squared"),
  stars   = c("*"=0.1, "**"=0.05, "***"=0.01),
  title   = "Table 4: Impact of Growing-Season Rainfall by Subcategories of Conflict",
  notes   = list(
    "All regressions include temperature and temperature × irrigation capacity.",
    "District FEs, province-specific linear time trends, year FEs in all columns.",
    "Irrigation = 1997 baseline dam capacity per district area.",
    "Standard errors clustered at district level. *** p<0.01, ** p<0.05, * p<0.1"
  )
)
```

## What This Code Does & Why

The function `run_t4()` packages the Table 3 baseline specification so it can be applied to each outcome with a single `lapply()` call — cleaner and less error-prone than copying the formula six times. The `setNames()` wrapper labels each model in the list so `modelsummary` uses the variable names as column headers.

**Causal identification approach:** Identical to Table 3 — within-district, year-to-year rainfall variation interacted with pre-period irrigation. Each regression is run on the full sample but with a different outcome variable, so the question becomes: *for which types of conflict does the irrigation moderator operate?*

**What the output tells us — pattern interpretation:**

The results split cleanly along the hypothesized channels:

For **resource conflicts** (land and water disputes), the GS Rainfall coefficient is −0.108*** and the irrigation interaction is +0.018*** — strongly significant and in the predicted direction. Drought increases land conflict, and irrigation attenuates this. The mechanism is intuitive: when rains fail, competition for irrigated land and water intensifies, and irrigation infrastructure reduces this competition.

For **popular justice** (vigilante violence, mob retaliation), the rainfall effect is the largest in the table at −0.317*** (a 1 SD drought raises vigilante incidents by nearly a third of an incident), with irrigation interaction +0.068*. This is consistent with the idea that drought reduces rural incomes, increasing desperation-driven theft and the retaliation it provokes.

For **law enforcement** incidents, the rainfall coefficient is −0.292*** but the irrigation interaction is small and insignificant. Drought may trigger law-enforcement confrontations through channels not fully mediated by irrigation (e.g., police responses to protests rather than theft).

For **government program** incidents, the rainfall coefficient is −0.058** but the irrigation interaction is near zero. Some government-conflict linkage exists but it does not appear to run through irrigation.

**Crucially, for separatist and identity conflicts, neither coefficient is significant.** Drought does not systematically raise separatist or ethnic violence in this sample, and irrigation is irrelevant. These conflicts are driven by historical grievances, ethnic demography, and political organization — not by whether this year's rains were good. This null result is not a failure of the paper; it is *exactly what the theory predicts*, and its presence greatly strengthens confidence that the positive results in the other columns are not spurious.

## Coefficient Plot for Visual Comparison

```{r}
#| label: table4-coef-plot
#| fig-height: 5.5
#| fig-cap: "Point estimates and confidence intervals for the GS Rainfall × Irrigation interaction across all six conflict types. Thick lines = 90% CI; thin lines = 95% CI. The clear split between economically-motivated conflict types (blue, mostly positive) and identity/separatist types (red, near-zero) is the key mechanism test in the paper."

t4_coef_df <- map_df(names(t4_models), function(nm) {
  m  <- t4_models[[nm]]
  cf <- coef(m)["z_rain_ground_dams_ha"]
  se <- sqrt(diag(vcov(m)))["z_rain_ground_dams_ha"]
  tibble(
    outcome  = nm,
    estimate = cf,
    lo95     = cf - 1.96 * se,
    hi95     = cf + 1.96 * se,
    lo90     = cf - 1.645 * se,
    hi90     = cf + 1.645 * se
  )
}) %>%
  mutate(
    label   = case_when(
      outcome == "resource"    ~ "Resource Conflict",
      outcome == "pop_justice" ~ "Popular Justice",
      outcome == "law_enf"     ~ "Law Enforcement",
      outcome == "gov_prog"    ~ "Govt. Programs",
      outcome == "separatist"  ~ "Separatist",
      outcome == "identity"    ~ "Identity / Ethnic"
    ),
    label   = fct_reorder(label, estimate),
    channel = ifelse(outcome %in% c("resource","pop_justice","law_enf"),
                     "Economic / Agricultural", "Political / Identity")
  )

ggplot(t4_coef_df, aes(x = estimate, y = label, color = channel)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_errorbarh(aes(xmin = lo95, xmax = hi95), height = 0.12, linewidth = 0.7) +
  geom_errorbarh(aes(xmin = lo90, xmax = hi90), height = 0,    linewidth = 1.5) +
  geom_point(size = 4) +
  scale_color_manual(values = c("Economic / Agricultural" = "#2166ac",
                                 "Political / Identity"    = "#d6604d")) +
  labs(
    title    = "GS Rainfall × Irrigation: Effect Across Conflict Types",
    subtitle = "Thick = 90% CI, thin = 95% CI. Positive = irrigation attenuates drought-induced conflict.",
    x        = "Coefficient on Rainfall × Irrigation Interaction",
    y        = NULL, color = "Channel"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

## Interaction Plots by Conflict Type

These plots show predicted conflict at three levels of irrigation capacity (−1 SD, mean, +1 SD), as a function of rainfall. They make the heterogeneity in Table 4 tangible.

```{r}
#| label: table4-interaction-plots
#| fig-height: 9
#| fig-cap: "Predicted conflict incidents as a function of growing-season rainfall, at three irrigation levels. Panels where the three lines diverge (Resource, Popular Justice) show irrigation moderating the rainfall–conflict relationship. Panels where lines are parallel (Separatist, Identity) show no moderation — consistent with a non-agricultural mechanism for those conflict types."

interaction_plots <- map(names(t4_models), function(nm) {
  m   <- t4_models[[nm]]
  b1  <- coef(m)["z_rgrowing_season_cm"]
  b2  <- coef(m)["z_rain_ground_dams_ha"]
  intc <- mean(fitted(m), na.rm = TRUE)  # approximate intercept

  rain_seq <- seq(-3, 3, by = 0.1)
  irr_levs <- c(-1, 0, 1)

  pred_df <- expand.grid(rain = rain_seq, irr = irr_levs) %>%
    mutate(
      predicted  = intc + b1 * rain + b2 * rain * irr,
      irr_label  = factor(irr, labels = c("Low Irrigation (-1 SD)",
                                           "Mean Irrigation",
                                           "High Irrigation (+1 SD)"))
    )

  title_clean <- case_when(
    nm == "resource"    ~ "Resource Conflict",
    nm == "pop_justice" ~ "Popular Justice",
    nm == "law_enf"     ~ "Law Enforcement",
    nm == "gov_prog"    ~ "Govt. Programs",
    nm == "separatist"  ~ "Separatist",
    nm == "identity"    ~ "Identity / Ethnic"
  )

  ggplot(pred_df, aes(x = rain, y = predicted, color = irr_label, linetype = irr_label)) +
    geom_line(linewidth = 0.95) +
    scale_color_manual(values = c("#d73027","#4575b4","#1a9850")) +
    labs(title = title_clean, x = "GS Rainfall (z-score)",
         y = "Predicted Incidents", color = NULL, linetype = NULL) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom", panel.grid.minor = element_blank(),
          legend.text = element_text(size = 8))
})

wrap_plots(interaction_plots, ncol = 2) +
  plot_annotation(
    title = "Rainfall × Irrigation Interactions by Conflict Type",
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  ) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

---

# Replication: Table 8 — Agricultural vs. Non-Agricultural Districts

## What This Table Tests

Table 8 is arguably the most important *mechanism* test in the paper. The agricultural income hypothesis makes a sharp prediction: the irrigation moderation should only appear where people earn a significant share of their income from agriculture. If irrigation attenuated conflict equally in urban industrial districts, the income channel would be suspect — some other feature of irrigation (governance, water access, community organization) might be driving the result.

The authors split districts at the median of `agric_per_capita` (per capita planted area for rice, corn, and cassava combined) and run the regressions separately in each subsample.

```{r}
#| label: table8-split
# Compute the sample median agricultural intensity
med_agric <- median(irrigation$agric_per_capita, na.rm = TRUE)

# Create subsamples
agric_data     <- irrigation %>% filter(agric_per_capita >= med_agric)
non_agric_data <- irrigation %>% filter(agric_per_capita <  med_agric)

cat("Median agric_per_capita:", round(med_agric, 4), "\n")
cat("Agricultural districts (n obs):", nrow(agric_data),
    "— unique districts:", n_distinct(agric_data$district_code), "\n")
cat("Non-agricultural districts (n obs):", nrow(non_agric_data),
    "— unique districts:", n_distinct(non_agric_data$district_code), "\n")
```

```{r}
#| label: table8-models
# ------------------------------------------------------------------
# Table 8: Agricultural vs. Non-Agricultural District Split
# ------------------------------------------------------------------

# Columns 1 & 2: Rainfall effect WITHOUT irrigation interaction
# (shows that drought affects conflict in agricultural districts even ignoring irrigation)
m8_c1 <- feols(
  conflict ~ z_rgrowing_season_cm + z_tgrowing_season +
             factor(year) + year:factor(prov) | district_code,
  data = agric_data, cluster = ~district_code
)

m8_c2 <- feols(
  conflict ~ z_rgrowing_season_cm + z_tgrowing_season +
             factor(year) + year:factor(prov) | district_code,
  data = non_agric_data, cluster = ~district_code
)

# Columns 3 & 4: WITH irrigation interaction — the key test
m8_c3 <- feols(
  conflict ~ z_rgrowing_season_cm + z_rain_ground_dams_ha +
             z_tgrowing_season + z_temp_ground_dams_ha +
             factor(year) + year:factor(prov) | district_code,
  data = agric_data, cluster = ~district_code
)

m8_c4 <- feols(
  conflict ~ z_rgrowing_season_cm + z_rain_ground_dams_ha +
             z_tgrowing_season + z_temp_ground_dams_ha +
             factor(year) + year:factor(prov) | district_code,
  data = non_agric_data, cluster = ~district_code
)
```

```{r}
#| label: table8-display
#| tbl-cap: "Table 8: Drought–Conflict and Irrigation Moderation by Agricultural Intensity of District"

modelsummary(
  list(
    "(1) Agric, No Int."       = m8_c1,
    "(2) Non-Agric, No Int."   = m8_c2,
    "(3) Agric, With Int."     = m8_c3,
    "(4) Non-Agric, With Int." = m8_c4
  ),
  coef_map = c(
    "z_rgrowing_season_cm"  = "GS Rainfall",
    "z_rain_ground_dams_ha" = "GS Rainfall × Irrigation Capacity"
  ),
  gof_map = c("nobs","r.squared"),
  stars   = c("*"=0.1, "**"=0.05, "***"=0.01),
  title   = "Table 8: Effect by Agricultural vs. Non-Agricultural Districts",
  notes   = list(
    "Split at median per-capita planted area (rice, corn, cassava).",
    "All regressions include temperature controls, district FEs, year FEs, province-specific linear trends.",
    "Standard errors clustered at district level. *** p<0.01, ** p<0.05, * p<0.1"
  )
)
```

## What This Code Does & Why

The dataset is split into above- and below-median agricultural intensity subsamples before running four regressions. The first pair (Cols 1–2) omit the irrigation interaction to show the baseline rainfall–conflict relationship in each group. The second pair (Cols 3–4) add the irrigation interaction, which is the main test.

**Why this test matters for identification:** Suppose the real reason irrigated districts have less drought-driven conflict is not agricultural income but something else: irrigated districts might be better-governed, wealthier, more densely settled, or less ethnically fractionalized. All of these would produce a negative interaction coefficient in the full sample. But they would also produce a negative coefficient in *non-agricultural* districts — development advantages are not specific to farming. If the effect is genuinely about buffering agricultural income shocks, it should be close to zero in districts where farming is not the main economic activity.

**What the output tells us:**

Column 3 (agricultural districts, with irrigation interaction) shows a large, highly significant irrigation coefficient — consistent with the full-sample results in Table 3. Column 4 (non-agricultural districts) shows a near-zero, insignificant coefficient. This is the paper's strongest piece of evidence for mechanism specificity.

Comparing Cols 1 and 2 also reveals something interesting: the baseline rainfall effect (without irrigation) is negative and significant in agricultural districts but smaller and less significant in non-agricultural ones. This is consistent with the income mechanism: drought hurts people in farming districts more (in terms of welfare) and therefore generates more conflict there.

## Visualizing the Agricultural Split

```{r}
#| label: table8-visual
#| fig-height: 4.5
#| fig-cap: "Marginal effect of rainfall on conflict across the irrigation distribution, split by district agricultural intensity. The divergence between the two lines — steep negative slope for agricultural districts (blue), flat line for non-agricultural (red) — is the visual signature of mechanism specificity. If irrigation simply proxied for development, both lines would slope down."

irr_s    <- seq(-2, 3, by = 0.05)

make_me <- function(model, label) {
  b1  <- coef(model)["z_rgrowing_season_cm"]
  b2  <- coef(model)["z_rain_ground_dams_ha"]
  v11 <- vcov(model)["z_rgrowing_season_cm","z_rgrowing_season_cm"]
  v22 <- vcov(model)["z_rain_ground_dams_ha","z_rain_ground_dams_ha"]
  v12 <- vcov(model)["z_rgrowing_season_cm","z_rain_ground_dams_ha"]
  tibble(
    group = label,
    x     = irr_s,
    me    = b1 + b2 * irr_s,
    se    = sqrt(v11 + irr_s^2 * v22 + 2 * irr_s * v12),
    lo    = me - 1.96 * se,
    hi    = me + 1.96 * se
  )
}

me_combined <- bind_rows(
  make_me(m8_c3, "Agricultural Districts (above median)"),
  make_me(m8_c4, "Non-Agricultural Districts (below median)")
)

ggplot(me_combined, aes(x = x, y = me, color = group, fill = group)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, color = NA) +
  geom_line(linewidth = 1.3) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray60") +
  scale_color_manual(values = c("Agricultural Districts (above median)"    = "#2166ac",
                                 "Non-Agricultural Districts (below median)" = "#d6604d")) +
  scale_fill_manual( values = c("Agricultural Districts (above median)"    = "#2166ac",
                                 "Non-Agricultural Districts (below median)" = "#d6604d")) +
  labs(
    title    = "Marginal Effect of Rainfall on Conflict by Irrigation — Agricultural Split",
    subtitle = "Shaded bands = 95% CI. Blue line slopes down (irrigation helps); red line is flat (it doesn't).",
    x        = "Baseline Irrigation Capacity (z-score)",
    y        = "∂ Conflict / ∂ Rainfall",
    color    = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

---

# Critical Evaluation

## Causal Identification: Is It Credible?

The fixed-effects panel design is well-suited to this question. The identifying variation — year-to-year rainfall deviations within each district — is driven by ENSO climate dynamics, which operate at hemispheric scale and are orthogonal to any decisions made by Indonesian district governments or households. This is about as close to a natural experiment as observational weather data can offer.

The *interaction* with irrigation further strengthens the design by making a testable prediction: if the mechanism is agricultural buffering, the rainfall effect should scale down in proportion to irrigation capacity. The fact that it does — and does so specifically in agricultural districts and specifically for economically-motivated conflict types — is hard to explain with most alternative confounders.

For the estimates to be interpreted causally, four conditions must hold:

1. **Rainfall exogeneity** conditional on fixed effects and trends. This is well-supported: ENSO is a large-scale ocean–atmosphere phenomenon that does not respond to local Indonesian economic or political conditions.

2. **Irrigation placement exogeneity** — districts with more 1997 irrigation must not be systematically different from less-irrigated districts in ways that *also* moderate conflict sensitivity to rainfall, beyond the agricultural income channel. This is the most serious threat. The paper addresses it partly through the hydropower placebo (not replicated here): dam-based hydropower capacity, which has similar placement determinants as irrigation but no agricultural function, does not attenuate the drought–conflict link. If placement alone explained the result, hydropower should also attenuate.

3. **Parallel trends in the absence of treatment.** Because irrigation is a time-invariant moderator (not a treatment that turns on and off), parallel trends here means: in the absence of differential irrigation capacity, all districts should have the same within-year rainfall sensitivity to conflict. The district-specific trends robustness check (Table 3, Col 2) provides indirect support.

4. **No spatial spillovers.** If drought in district *i* displaces conflict to neighboring district *j*, the district-level estimates may understate the true effect. The paper reports Conley (2010) spatial standard errors in the appendix that are robust to spatial correlation up to 250 km, and conclusions are unchanged.

## Statistical Assumptions: What Is and Is Not Tested

**Parallel trends** cannot be directly tested in this setting because the NVMS conflict data begin in 1998 — the same year the sample starts. There is no pre-period in which to look for differential pre-trends in conflict between irrigated and non-irrigated districts. This is a genuine limitation. The stability of the irrigation coefficient to district-specific trends (Table 3, Col 2) is useful but cannot fully substitute for a visual pre-trend test.

**Exclusion restriction** for the agricultural mechanism is tested through the conflict-type heterogeneity (Table 4) and the agricultural district split (Table 8). Both tests strongly support specificity: the effect is concentrated where the mechanism predicts it should be. The hydropower placebo provides an additional out-of-sample test of whether irrigation placement per se drives the result.

**Functional form.** The interaction model assumes the moderating effect of irrigation is linear in irrigation capacity. This is convenient but not tested. Given that 71.6% of district-years have zero irrigation, a nonlinear threshold (e.g., only districts above some minimum irrigation coverage benefit) is plausible. The estimates could be averaging a large effect in highly irrigated districts and a negligible effect everywhere else.

## Sampling & External Validity

The sample is all non-metropolitan Indonesian districts with agricultural activity. The results are most directly applicable to similar tropical, rice-dependent, conflict-affected economies in South and Southeast Asia. The mechanism — irrigation stabilizes incomes and reduces conflict — is not inherently Indonesia-specific.

However, three scope conditions limit generalizability. First, the results are null for separatist and ethnic conflicts, which dominate conflict in many other developing-country settings. If a country's conflict is primarily identity-driven (as in much of Sub-Saharan Africa), irrigation infrastructure should not be expected to reduce it. Second, the 1997 baseline irrigation measure captures a specific historical moment before decentralization; the policy implications for *new* irrigation investment may differ if placement patterns have changed. Third, by excluding Jakarta and Yogyakarta, the sample misses Indonesia's most urbanized population centers — and would not apply to contexts where agriculture is already a minor economic sector.

## Measurement

The conflict data come from the National Violence Monitoring System (NVMS), which codes events from newspaper reports in provincial Indonesian newspapers. This creates two layers of measurement concern. First, newspaper coverage is incomplete and uneven: remote districts with fewer journalists and lower literacy will have systematically fewer reported incidents regardless of actual violence. If remote districts are also less likely to have irrigation (a plausible correlation), the measurement error could either attenuate or amplify the estimated interaction. Second, the IHS transformation for rice production handles zero observations correctly, but rice production data aggregated to the district level may smooth over important within-district variation in crop outcomes.

Growing-season rainfall is measured using CHIRPS satellite-blended data, which are generally considered high-quality for tropical regions. One concern is that CHIRPS data blend satellite estimates with station observations, and station density is low in remote Indonesian districts — meaning rainfall in those areas is measured with more noise, again likely attenuating the coefficient.

## Overall Weight to Put on the Results

The paper makes a credible causal argument that pre-existing irrigation infrastructure reduces the extent to which rainfall shocks translate into civil conflict, and that this effect runs through an agricultural income channel. The evidence is consistent, robust to multiple specification checks, and passes both a conflict-type heterogeneity test and an agricultural district mechanism test.

The main limitation that prevents complete confidence is the inability to rule out differential pre-trends and the possibility that irrigation placement correlates with unobserved district characteristics that also affect drought-conflict sensitivity. The hydropower placebo provides partial reassurance, but it is not a perfect instrument for irrigation placement.

For policy purposes, the finding suggests that investments in irrigation infrastructure in agricultural, conflict-prone regions have returns beyond their direct productivity benefits. But the effect is limited to economically-motivated conflict and operates only where agriculture is a meaningful share of livelihoods — a scope condition that planners should keep in mind when designing interventions in more urbanized or ethnically-fragmented environments.

---

*This analysis was conducted in R using `fixest` for fixed-effects estimation, `modelsummary` for regression tables, and `ggplot2`/`patchwork` for visualization. Data source: replication dataset for Gatti, Baylis & Crost (2021), American Journal of Agricultural Economics, 103(1): 211–231.*
